---
title: 'Modeling Vegetation Stem Density from Emprical Data'
author: "Melanie Davis"  
date: "November 27, 2017"  
output:
 github_document
---

```{r include=FALSE}
	rm(list = ls())
	library(ggplot2)
	library(bbmle)
	library(nlstools)
	library(propagate)
	Veg=read.csv("E:\\UW PHD\\Dissertation\\Chapter_1\\CH1_Data\\CH1_Vegetation.csv")
```

##Data Preparation
  
The data input is a file containing stem density values for five years of vegetation surveys.
Column 1 (ID) is sample ID including year, month, site, and transect, Column 2 (Density) is 
the stem density of the 0.5 x 0.5 m quadrat, Column 3 (Chan_Dist_m) is the distance from 
the edge of the nearest channel in meters, Column 4 (Elev_m) is the quadrat elevation in meters NAVD88,
Column 5 (Salinity) is the measured soil pore salinity at the quadrat.
  
```{r, echo=FALSE}
	head(Veg)
```
  
First we need to calcualte inundation duration using the elevation measurements. Inundation duration 
is the proportion of time each cell of elevation x meters will be covered by the tide.  
  
```{r}
Veg$Inundation = 1/(1 + (0.033*exp(-0.9*1.34))*exp(0.9*Veg$Elev_m))^(1/0.06)
plot(Veg$Elev_m, Veg$Inundation, pch=19)
```   
  
Now we look at the distribution of the stem density data and run some preliminary plots.

```{r, echo=FALSE}
	hist(Veg$Density)
	hist(log(Veg$Density + 1))
	hist(sqrt(Veg$Density))
```
  
The data are zero-skewed poisson (count) data. Log-transforming or square root transformation helps,
except for zero-skewness.
  
```{r, echo=FALSE}
	plot(Veg$Inundation, Veg$Density)
	abline(lm(Veg$Density~Veg$Inundation))
```

Stem density appears to follow a gaussian function with respect to inundation

```{r, echo=FALSE}
	plot(Veg$Salinity, Veg$Density)
	abline(lm(Veg$Density~Veg$Salinity))
```
  
Stem density appears to decrease with increasing salinity.  


```{r, echo=FALSE}
	plot(Veg$Chan_Dist_m, Veg$Density)
	abline(lm(Veg$Density~Veg$Chan_Dist_m))
```
  
There doesn't appear to be a realtionship between stem density and distance to channel edge.  
  
Now that we have evaluated the relationship between soil pore salinity and inundation duration, we can
run the model parameterization and selection procedure.
  
##Model Parameterization

First let's run some linear models to parse out significant drivers. 
  
```{r, warning=FALSE}
	VEGlm1 = lm(Density ~ Inundation, data=Veg)
	VEGlm2 = lm(Density ~ Salinity, data=Veg)
	VEGlm3 = lm(Density ~ Inundation + Salinity, data=Veg)
	VEGlm4 = lm(Density ~ Inundation*Salinity, data=Veg)
	VEGlm5 = lm(log(Density+1) ~ Inundation, data=Veg)
	VEGlm6 = lm(log(Density+1) ~ Salinity, data=Veg)
	VEGlm7 = lm(log(Density+1) ~ Inundation + Salinity, data=Veg)
	VEGlm8 = lm(log(Density+1) ~ Inundation*Salinity, data=Veg)
	VEGlm9 = lm(sqrt(Density) ~ Inundation, data=Veg)
	VEGlm10 = lm(sqrt(Density) ~ Salinity, data=Veg)
	VEGlm11 = lm(sqrt(Density) ~ Inundation + Salinity, data=Veg)
	VEGlm12 = lm(sqrt(Density) ~ Inundation*Salinity, data=Veg)
	
	AIC(VEGlm1,VEGlm2,VEGlm3,VEGlm4,VEGlm5,VEGlm6,VEGlm7,VEGlm8,VEGlm9,VEGlm10,VEGlm11,VEGlm12)
	as.array(c(sigma(VEGlm1),sigma(VEGlm2),sigma(VEGlm3),sigma(VEGlm4),sigma(VEGlm5),
		sigma(VEGlm6),sigma(VEGlm7),sigma(VEGlm8)))

summary(VEGlm7)
```
  
The additive model of Inundation and Salinity appears to be the best-fit model, although salinity is not significant.
  
As we observed above, the relationship between stem density and inundation duration appeared to be Gaussian, with
lower stem densities at the longest and shortest inundation durations. The Gaussian function is:
  
$V_x = \alpha e^{\frac{-(I_x-\beta)^2}{2\gamma^2}}$ 
  
where $$\alpha = \frac{1}{sd}\sqrt{\frac{2}{\pi}}$$ $$\beta = mean$$ and $$\gamma = sd$$  
  
Stem density also appears to decline with increasing salinity, and because this relationship must have
a lower asymptote, an exponential decay function may serve as the best representation. 
  
$V_x = Ne^{-\lambda S_x}$ 
  
We can apply this equation to the Inundation and Salinity variables using the "nls" function.

```{r}
VEG_func1 = function(alpha, beta, gamma){
Stems = alpha*exp((-(Veg$Inundation - beta)^2)/(2*(gamma^2)))
}

VEGmle1=nls(Density ~ alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
data=Veg, 
start = list(alpha=100, beta=0.15, gamma=0.2),
lower=c(0, 0, 0),
upper=c(10000, 1, 1),
algorithm="port")
summary(VEGmle1)
AIC(VEGlm1, VEGmle1)
sigma(VEGmle1)
```
  
As you can see, the Gaussian model offers a much better fit. We can run a model selection procedure 
using multiple, interactive combinations of the above-tested variables:

```{r, echo=FALSE, warning=FALSE}
VEGmle2=nls(Density ~ s1 + s2*Salinity + alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
data=Veg, 
start = list(s1 = 0, s2 = 2, alpha=100, beta=0.15, gamma=0.2),
lower=c(-100, 0, 0, 0, 0),
upper=c(100, 1000, 10000, 1, 1),
algorithm="port")
summary(VEGmle2)

		VEGmle3=nls(Density ~ (alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2)))) + (s1*Salinity*exp((-(Inundation - beta)^2)/(2*(gamma^2)))), 
			data=Veg, 
			start = list(alpha = 5, s1 = 1, beta=0.15, gamma=0.2),
			lower=c(0, 0, 0, 0),
			upper=c(10000, 100, 1, 1),
			algorithm="port")
		summary(VEGmle3)

		VEGmle4=nls(Density ~ alpha*Salinity*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(alpha = 10, beta=0.15, gamma=0.2),
			lower=c(0, 0, 0),
			upper=c(1000, 1, 1),
			algorithm="port")
		summary(VEGmle4)

		VEGmle5=nls(sqrt(Density) ~ alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(alpha=10, beta=0.15, gamma=0.2),
			lower=c(0, 0, 0),
			upper=c(10000, 1, 1),
			algorithm="port")
		summary(VEGmle5)

		VEGmle6=nls(sqrt(Density) ~ s1 + s2*Salinity + alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(s1 = 0, s2 = 2, alpha=100, beta=0.15, gamma=0.2),
			lower=c(-100, 0, 0, 0, 0),
			upper=c(100, 1000, 10000, 1, 1),
			algorithm="port")
		summary(VEGmle6)

		VEGmle7=nls(sqrt(Density) ~ (alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2)))) + (s1*Salinity*exp((-(Inundation - beta)^2)/(2*(gamma^2)))), 
			data=Veg, 
			start = list(alpha = 5, s1 = 1, beta=0.15, gamma=0.2),
			lower=c(0, 0, 0, 0),
			upper=c(500, 100, 1, 1),
			algorithm="port")
		summary(VEGmle7)

		VEGmle8=nls(sqrt(Density) ~ alpha*Salinity*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(alpha = 10, beta=0.15, gamma=0.2),
			lower=c(0, 0, 0),
			upper=c(1000, 1, 1),
			algorithm="port")
		summary(VEGmle8)

		VEGmle9=nls(Density ~ N*exp(-lambda*Salinity), 
			data=Veg, 
			start = list(N = 3000, lambda=0),
			lower=c(0, -5),
			upper=c(10000, 5),
			algorithm="port")
		summary(VEGmle9)

		VEGmle10=nls(Density ~ N*exp(-lambda*Salinity) + alpha*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(N = 3000, lambda=0, alpha=100, beta=0.12, gamma=0.11),
			lower=c(0, -5, 0, 0, 0),
			upper=c(10000, 5, 10000, 100, 100),
			algorithm="port")
		summary(VEGmle10)

		VEGmle11=nls(Density ~ N*exp(-lambda*Salinity)*exp((-(Inundation - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(N = 3000, lambda=0, beta=0.12, gamma=0.11),
			lower=c(0, -5, 0, 0),
			upper=c(10000, 5, 100, 100),
			algorithm="port")
		summary(VEGmle11)

		VEGmle12=nls(Density ~ alpha*exp((-(Salinity - beta)^2)/(2*(gamma^2))), 
			data=Veg, 
			start = list(alpha = 100, beta=3, gamma=3),
			lower=c(0, 0, 0),
			upper=c(10000, 1000, 1000),
			algorithm="port")
		summary(VEGmle12)

		VEGmle13=nls(Density ~ alpha*exp((-(Salinity - beta_s)^2)/(2*(gamma_s^2)))*exp((-(Inundation - beta_i)^2)/(2*(gamma_i^2))), 
			data=Veg, 
			start = list(alpha = 100, beta_s=3, gamma_s=3, beta_i=0.1, gamma_i=0.1),
			lower=c(0, 0, 0, 0, 0),
			upper=c(10000, 1000, 1000, 100, 100),
			algorithm="port")
		summary(VEGmle13)

AIC(VEGmle1,VEGmle2,VEGmle3,VEGmle4,VEGmle5,VEGmle6,VEGmle7,VEGmle8,VEGmle9,VEGmle10,VEGmle11,VEGmle12,VEGmle13)
```
The best-fit model here is the interactive model between Inundation Duration (Gaussian) and Soil Pore Salinity
(Exp Decay).
  
The modeled output looks like this:  

```{r include=FALSE}
Inundation_test = seq(0,0.75,0.005)
Salinity_test = array(mean(Veg$Salinity, na.rm=TRUE),length(Inundation_test))
newdat=as.data.frame(cbind(Inundation_test, Salinity_test))
colnames(newdat) = c("Inundation","Salinity")

MLE_pred = predictNLS(VEGmle11, newdata=data.frame(Salinity=20, Inundation=Inundation_test), interval="prediction", alpha=0.05)
	MLE_pred_out=as.data.frame(MLE_pred[1])
	head(MLE_pred_out)

	SubVEG = subset(Veg, !Salinity > 30)
	SubVEG = subset(SubVEG, !Salinity < 10)	
``` 
```{r echo=FALSE}
ggplot() +
	geom_area(aes(x=Inundation_test, y=(MLE_pred_out$summary.Sim.97.5.)), fill="lightgray") +
	geom_area(aes(x=Inundation_test, y=(MLE_pred_out$summary.Sim.2.5.)), fill="white") +
	geom_line(aes(x=Inundation_test, y=(MLE_pred_out$summary.Sim.Mean)), lwd=1.5) +
	geom_point(aes(x=Veg$Inundation, y=Veg$Density), pch=19, col="black") +
	geom_point(aes(x=SubVEG$Inundation, y=SubVEG$Density), pch=19, col="red") +
	scale_x_continuous(limits=c(0,0.75)) +
	scale_y_continuous(limits=c(0,4000)) +
	xlab("Inundation Duration (%)") +
	ylab("Stem Density (m-2)") +
	ggtitle("Salinity = 20 PSU") +
	theme_classic()+
	theme(plot.title = element_text(hjust = 0.5))
```